---
title: "Answers to SQL Code Questions"
author: "Michael Boerman"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    code_download: true
---

<base target="_top"/>

Want to view this file? [See it on RPubs](https://rpubs.com/michaelboerman/experian_interview_sql_questions).

Want to run this file or make local edits? [Open it on Posit Cloud](https://posit.cloud/content/5828225).

Want to view the change history for this file? [View it on GitHub](https://github.com/michaelboerman/experian_takehome/blob/main/code/sql_questions.Rmd).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,           # show the code by default (will be hidden by code folding)
  results = 'show',         # do print the results of each chunk
  message = FALSE           # don't print messages - but do print errors/warnings.
)

library(readxl)
library(here)
library(tidyr)
library(dplyr)
library(naniar) # for replace_with_na function :) 

# let's parameterize this in case it were to change
# or we wanted to run the code on a different, similar file.
excel_file_name <- "Interview Assessment Workbook 04.18.23 v.5.xlsx"
```

# Reading the Data
Let's start by reading in the data and viewing it for potential issues. 
```{r}
raw_app_data <- here("input_data", excel_file_name) |> 
  read_xlsx(sheet = "Application")

raw_app_data
```
I see values in `Loan Booked` and `Loan Performance` with "n/a" as characters, as well as the last two rows having ".".

I also see that `Loan Amount` column is listed as a character vector, but it should be a numeric. 
This helped me find a period in row 549 and an "n/a" in row 588.

I'll replace the "n/a" and "." values with R's `NA` value using `naniar::replace_with_na_all`.
I could also add arguments to the original `read_xlsx` call to specify na strings and column types; for example, `read_xlsx(filepath, sheet = "Application", na = c("n/a", "."))`. This would also read the Loan Amount column type correctly as numeric instead of character. Nonetheless, for exploratory sake, I make these changes after reading the raw data:

```{r}
# replace_with_na_all is a function from naniar similar to running a case_when
# or running mutate(across(select(everything()), tidyr::na_if("n/a")))
cleaned_app_data <- raw_app_data |> 
  replace_with_na_all(condition = ~.x %in% c("n/a", ".")) |> 
  
  # now that the character strings are replaced with NA,
  # we can convert this from character to numeric
  mutate(`Loan Amount`  = as.numeric(`Loan Amount`))

cleaned_app_data
```

Let the analysis begin!