---
title: "Answers to SQL Code Questions"
author: "Michael Boerman"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    code_download: true
---

<base target="_top"/>

This file can be [viewed on RPubs](https://rpubs.com/michaelboerman/experian_interview_sql_questions), or [run on Posit Cloud](https://posit.cloud/content/5828225), and it [lives on GitHub here](https://github.com/michaelboerman/experian_takehome/blob/main/code/sql_questions.Rmd).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  results = 'hide',
  message = FALSE
)
```


```{r packages_connection_setup}
library(dplyr)
library(RSQLite)
library(DBI)
library(here)
library(dbplyr)
library(readr)
library(stringr)
library(purrr)

# Create a local database in memory
# This will be removed/reset each time the R session is terminated.
con <- dbplyr::src_memdb()$con
```


```{r create_tables_function}

# We need a function to read in the SQL files and parse out each query within the file.
# Generally, a file has three commands:
# 1. drop the table if it already exists,
# 2. create the table,
# 3. insert values into the table.
# 
# each of these commands will become a character vector in the list
parse_sql_file <- function(sql_file) { 
  
  # test value
  # sql_file <- here("code/create_authors_table.sql")
  
  order_of_commands <- c("drop_table", "create_table", "insert_values")

  
  # Read in the file
  each_query_in_file  <- read_file(sql_file) |> 
    
    # split the whole string into sub strings before each ;
    str_split(";") |> 
    
    # remove the new lines and tab symbols.
    map(~ str_replace_all(.x, "\n", "")) |> 
    map(~ str_replace_all(.x, "\t", "")) |> 
    
    # convert from list of character vectors to just character vectors
    unlist()
  
  # since str_split will create a blank after the last query, remove this from the list
  # # don't look at the gawd-awful non-tidy code
  each_query_in_file <- each_query_in_file[-length(each_query_in_file)]
  
  # for easier viewing, give each character vector a descriptive name of the action performed
  # in its query.
  each_query_in_file <- each_query_in_file |> 
    set_names(order_of_commands) 
  
  return(each_query_in_file)
}


# This function executes any command.
# it accepts a single string as input.
execute_query <- function(query_string, con=con) {
  
  # test value
  # query_string <- parse_sql_file(here("code/create_authors_table.sql"))
  
  query_string |> 
    map(~ dbExecute(con, .x))
  
}


verify_tables_created <- function(table_names, con=con) {
  
  # verify that the list of table names provided at the top 
  # matches the list of sql table names stored in memory
  # after the code should have created them
  
  desired_tables <- table_names
  created_tables <- dbListTables(con)
  
  if (identical(desired_tables, created_tables) == FALSE) {
    for (table_to_have_been_created in desired_tables) {
      if (table_to_have_been_created %in% created_tables) {
        next()
      } else {
        print(str_glue("Warning! {table_to_have_been_created} was NOT created!"))
      }
    }
  }
}
```


```{r create_populate_tables}
# Now, execute the functions above.
# We need to use `purrr::map`, since we're iterating over a list.

# list of table names, which are also found in the file names
table_names <- c("authors", "books", "employees", "salaries")

paste0(here()) |> 
  paste0("/code/create_", table_names, "_table.sql") |> 
  set_names(table_names) |> 
  map(parse_sql_file) |> 
  map(~execute_query(.x, con))
```


## SQL Question 1
**Create a SQL query that shows the TOP 3 authors who sold the most books in total**
```{sql, connection = con, class.source = 'fold-show'}
select author_name, sum(sold_copies) as total_books_sold
  from authors
  left join books 
       on authors.book_name = books.book_name
 group by author_name
 order by sold_copies desc
```
P.S.: I've enjoyed many books from John Paulos, particularly his "A Mathematician Reads the Newspaper" and "A Mathematician Plays the Stock Market". He's got such a fun perspective on life -- except toward those who are not-so-mathematically-competent, as I found in "Innumeracy"! 
Upon seeing McDougall's copies sold as orders of magnitude more than Levitt, and a quick google search, I suspect data-bias by the table's creator - ha! 

## SQL Question 2
**Print every department where the average salary per employee is less then or equal to $700**
```{sql, connection = con, class.source = 'fold-show'}
-- instead of a nested-select clause,
-- use a common table expression 
-- to first find the average salaries for all departments
with dept_salaries as (
  select department_name, avg(salary) as average_salary
      from employees
      left join salaries
           on  employees.employee_id = salaries.employee_id
           and employees.employee_name = salaries.employee_name -- for safe measure!
     group by department_name
)

-- now we can filter the table above to just those <= 700 average salary.
select * 
  from dept_salaries
 where average_salary <= 700
```
